Tropical mathematics and the Lambda-calculus I

### 2023-07-29 CSL24/section7.tex
commit

The connections between differential $\lambda$-calculus (and differential linear logic), relational semantics, and non-idempotent intersection types are very well-studied (see \cite{decarvalho2018}, and more recently, \cite{Mazza2016} for a more abstract perspective, and \cite{Olimpieri2021, Galal2021} for a 2-categorical, or proof-relevant, extension).
As we said, the relational semantics over the tropical semi-ring was quickly explored in \cite{Manzo2013}, to provide a ``best case'' resource analysis of a $\B{PCF}$-like language with non-deterministic choice. 
\emph{Probabilistic coherent spaces} \cite{Ehrhard2011}, a variant of  the relational semantics, provide an interpretation of higher-order probabilistic programs
as analytic functions. In \cite{Ehrhard2022} it was observed that such functions satisfy a local Lipschitz condition somehow reminiscent of our examples in Section \ref{section4}.

@@

More generally, the literature on program metrics in denotational semantics is vast. Since at least \cite{VANBREUGEL20011} metric spaces, also in Lawvere's generalized sense \cite{Lawvere1973}, have been exploited as an alternative framework to standard, domain-theoretic, denotational semantics. 
While standard categories of metric spaces are not models of the full simply typed $\lambda$-calculus, several constructions of cartesian closed categories of metric spaces can be found in the literature. For instance, 
\emph{ultra}-metric spaces form a CCC, and have been shown to model $\B{PCF}$ \cite{Escardo1999}.
Also \emph{partial} metrics, introduced in \cite{matthews}, have been shown to provide models of $\STLC$, under suitable generalizations \cite{Geoffroy2020}.
More generally, \cite{Clementino2006} provides a general characterization of exponentiable objects in categories of (generalized) metric spaces, and \cite{PistoneLICS, PistoneFSCD2022} provide other ways to construct CCCs of
(generalized) metrics, including one based on locally Lipschitz maps, 
using ideas from {differential logical relations} and quantitative equational theories \cite{Plotk}.

@@

Finally, applications of tropical mathematics in computer science abound, originally for automata theory \cite{Chua1992, Simon}, and more recently
 in optimization methods for machine learning and other statistical models (see e.g.~\cite{Maragos2021, Pachter2004, Zhang2018}), optimization \cite{Akian2011, Akian2012}, and convex analysis \cite{Lucet2009}. While our discussion in Section \ref{section5} is inspired by a well-known application of tropical polynomials to Hidden Markov Models \cite{Pachter2004},  
the vast literature in this domain lets us think that other ways to 
apply tropical semantics to the analysis of higher-order programs might be studied.

%
%Other connections tropical/metric -> Fuji, ??
%Applications of tropical to computer science.
%Log-probabilities.
%Quantale-modules -> Abramski?
%
%

>>>>

The applications of tropical mathematics in computer science abound, e.g.~in automata theory \cite{Chua1992, Simon}, machine learning \cite{Maragos2021, Pachter2004, Zhang2018}, optimization \cite{Akian2011, Akian2012}, and convex analysis \cite{Lucet2009}.

The connections between differential $\lambda$-calculus (and differential linear logic), relational semantics, and non-idempotent intersection types are very well-studied (see \cite{decarvalho2018}, and more recently, \cite{Mazza2016} for a more abstract perspective, and \cite{Olimpieri2021, Galal2021} for a 2-categorical, or proof-relevant, extension).
As we said, the relational semantics over the tropical semiring was quickly explored in \cite{Manzo2013}, to provide a ``best case'' resource analysis of a $\mathrm{PCF}$-like language with non-deterministic choice. 
\emph{Probabilistic coherent spaces} \cite{Ehrhard2011}, a variant of  the relational semantics, provide an interpretation of higher-order probabilistic programs
as analytic functions. In \cite{Ehrhard2022} it was observed that such functions satisfy a local Lipschitz condition somehow reminiscent of our examples in Section \ref{section2}.

@@

More generally, the literature on program metrics in denotational semantics is vast. Since at least \cite{VANBREUGEL20011} metric spaces, also in Lawvere's generalized sense \cite{Lawvere1973}, have been exploited as an alternative framework to standard, domain-theoretic, denotational semantics, notably
\emph{ultra}-metrics and \emph{partial} metrics have been shown to form a CCC \cite{Escardo1999,PistoneLICS, PistoneFSCD2022}.

@@

%While our discussion in Section \ref{section5} is inspired by a well-known application of tropical polynomials to Hidden Markov Models \cite{Pachter2004},  
%the vast literature in this domain lets us think that other ways to 
%apply tropical semantics to the analysis of higher-order programs might be studied.
%
%
%Other connections tropical/metric -> Fuji, ??
%Applications of tropical to computer science.
%Log-probabilities.
%Quantale-modules -> Abramski?
%
%

### 2023-07-29 CSL24/section8.tex
commit

@@

While the first goal was here developed in detail, and at different levels of abstraction, for the second goal we only sketched a few interesting directions (best-case analysis, log-probabilities, differential privacy). We believe that exploring these ideas in more depth could be a fruitful direction; moreover, 
since both generalized metrics and quantale-modules have been largely studied in computer science, 
a natural question is if the generalized approach of Section \ref{section6} could  lead to new applications of metric and tropical methods to the $\lambda$-calculus.

Finally, since bounds on the Taylor expansion translate into Lipschitz conditions, 
 two interesting directions to explore are provided by (non-idempotent) intersection types and finiteness spaces \cite{Ehrhard2005}, as both methods are in principle capable of capturing \emph{finitary} bounds on the Taylor expansion. %
Notably, knowing that the application of a program $M$ to $N$ expands as a finite sum of linear applications may allow one to predict how sensitive $M$ will be ``around $N$''.

\newpage

>>>>

%{\color{red}due parole sul fatto che non sappiamo se si posso usare strumenti propri della tropical geo per lo studio dei programmi. Per esempio:}
%\begin{remark}\label{rem:troproots}
% Taking the usual $M$ as example, the $p\in[0,1]$ s.t.\ $(p,1-p)$ is a tropical root of $\trop Q_{\true}$ or of $\trop Q_{\false}$ provide, by definition, the values of the bias of $\oplus_p$ for which there are at least two different equiprobable paths of tossings from $M$ to its normal form.
% Moreover, looking at Equation~\ref{eq:max}, we see that the $p\in[0,1]$ s.t.\ $(-\log p,-\log(r-p))$ is a tropical root of, say, $\trop Q_{\true}$ are the values of the bias of $\oplus_p$ for which there are at least two different equiprobable occurrences of $\true$ that are smapled by $M$ during its tossings to normal form, knowing that some $\true$ was sampled.
%\end{remark}
%
%{\color{red}due parole generalised metric spaces and modules}
%
%
%{\color{red}due parole su PCoh, $\overline{R_{\geq 0}}$Rel e le loro tropicalizzazioni (gia accennato in \autoref{rmk:tropof01Rel})}

@@

While the first goal was here developed in detail, and at different levels of abstraction, for the second goal we only sketched a few interesting directions (best-case analysis, log-probabilities), and we leave their development to a second paper of this series. 
To conclude, while the main ideas of this article only use basic concepts from the toolbox of tropical mathematics, it looks like an exciting direction to look for applications of more advanced tools from tropical algebraic and differential geometry (e.g.~Newton polytopes, tropical varieties, tropical differential equations). 
%We believe that exploring these ideas in more depth could be a fruitful direction; moreover, 
%since both generalized metrics and quantale-modules have been largely studied in computer science, 
%a natural question is if the generalized approach of Section \ref{section6} could  lead to new applications of metric and tropical methods to the $\lambda$-calculus.

%Finally, since bounds on the Taylor expansion translate into Lipschitz conditions, 
% two interesting directions to explore are provided by (non-idempotent) intersection types and finiteness spaces \cite{Ehrhard2005}, as both methods are in principle capable of capturing \emph{finitary} bounds on the Taylor expansion. %
%Notably, knowing that the application of a program $M$ to $N$ expands as a finite sum of linear applications may allow one to predict how sensitive $M$ will be ``around $N$''.
